\documentclass[a4paper,UKenglish,cleveref, autoref,numberwithinsect]{lipics-v2019}

\title{Guarded Recursion in Agda via Sized Types}

\author{Niccol\`o Veltri}{Department of Computer Science, IT
  University of Copenhagen, Denmark}{nive@itu.dk}{https://orcid.org/0000-0002-7230-3436}{This work was supported by a research grant (13156) from VILLUM FONDEN,
  and by DFF-Research Project 1 Grant no. 4002-00442, from The Danish Council 
  for Independent Research for the Natural Sciences (FNU).}

\author{Niels van der Weide}{Institute for Computation and Information Sciences, Radboud University, Nijmegen, The Netherlands}{nweide@cs.ru.nl}{https://orcid.org/0000-0003-1146-4161}{}

\authorrunning{N. Veltri and N. van der Weide}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Niccol\`o Veltri, Niels van der Weide}

\ccsdesc[100]{Theory of Computation~Type Theory}
\ccsdesc[100]{Theory of Computation~Categorical Semantics}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm 

\keywords{guarded recursion, type theory, semantics, coinduction, sized types}%TODO mandatory; please add comma-separated list of keywords

\category{}%optional, e.g. invited paper

\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{We are thankful to Andreas Abel, Guillaume Allais, Herman Geuvers,
  Rasmus Ejlers M{\o}gelberg and Andrea Vezzosi for discussions and
valuable hints. We thank the anonymous referees for their useful comments.}%optional

\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as
%author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Herman Geuvers}
\EventNoEds{1}
\EventLongTitle{4th International Conference on Formal Structures for Computation and Deduction (FSCD 2019)}
\EventShortTitle{FSCD 2019}
\EventAcronym{FSCD}
\EventYear{2019}
\EventDate{June 24--30, 2019}
\EventLocation{Dortmund, Germany}
\EventLogo{}
\SeriesVolume{131}
\ArticleNo{33}

\usepackage{csquotes}
\usepackage[all,cmtip]{xy}

%%%%%%%%%%%%%   AMS MATH  %%%%%%%%%%%%%%
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{proof}
\usepackage{stmaryrd}

\usepackage[noinputencodingsetup]{agda}

% The following packages are needed because unicode
% is translated (using the next set of packages) to
% latex commands. You may need more packages if you
% use more unicode characters:

\usepackage{amssymb}
\usepackage{bbm}

\usepackage{newunicodechar}
\usepackage{stmaryrd}

% This handles the translation of unicode to latex:


% Some characters that are not automatically defined
% (you figure out by the latex compilation errors you get),
% and you need to define:

\newunicodechar{⊤}{$\top$}
\newunicodechar{≡}{$\equiv$}
\newunicodechar{₁}{$_1$}
\newunicodechar{₂}{$_2$}
\newunicodechar{₃}{$_3$}
\newunicodechar{Γ}{$\Gamma$}
\newunicodechar{Δ}{$\Delta$}
\newunicodechar{Σ}{$\Sigma$}
\newunicodechar{α}{$\alpha$}
\newunicodechar{β}{$\beta$}
\newunicodechar{η}{$\eta$}
\newunicodechar{γ}{$\gamma$}
\newunicodechar{ε}{$\varepsilon$}
\newunicodechar{θ}{$\theta$}
\newunicodechar{λ}{$\lambda$}
\newunicodechar{μ}{$\mu$}
\newunicodechar{κ}{$\kappa$}
\newunicodechar{π}{$\pi$}
\newunicodechar{∈}{$\in$}
\newunicodechar{⊞}{$\boxplus$}
\newunicodechar{⊠}{$\boxtimes$}
\newunicodechar{∁}{$\mathbb{C}$}
\newunicodechar{▸}{$\blacktriangleright$}
\newunicodechar{►}{$\blacktriangleright$}
\newunicodechar{▻}{$\vartriangleright$}
\newunicodechar{𝟙}{\textbf{1}}
\newunicodechar{∙}{$\cdot$}
\newunicodechar{∀}{$\forall$}
\newunicodechar{ℓ}{$\ell$}
\newunicodechar{∅}{$\varnothing$}
\newunicodechar{⟶}{$\rightarrow$}
\newunicodechar{□}{$\square$}
\newunicodechar{⇡}{$\uparrow$}
\newunicodechar{⊛}{$\circledast$}
\newunicodechar{⊎}{$\uplus$}
\newunicodechar{⟨}{$\langle$}
\newunicodechar{⟩}{$\rangle$}
\newunicodechar{■}{$\blacksquare$}
\newunicodechar{∎}{$\blacksquare$}
\newunicodechar{⇒}{$\Rightarrow$}
\newunicodechar{⊕}{$\oplus$}
\newunicodechar{⊗}{$\otimes$}
\newunicodechar{∼}{$\sim$}
\newunicodechar{≈}{$\approx$}
\newunicodechar{⊔}{$\sqcup$}
\newunicodechar{⟦}{$\llbracket$}
\newunicodechar{⟧}{$\rrbracket$}
\newunicodechar{⊥}{$\bot$}
\newunicodechar{∞}{$\infty$}
\newunicodechar{⇑}{$\Uparrow$}
\newunicodechar{≃}{$\simeq$}
\newunicodechar{∘}{$\circ$}
% Add more as you need them (shouldn't happen often).

\newcommand{\AD}[1]{\AgdaDatatype{#1}}
\newcommand{\AIC}[1]{\AgdaInductiveConstructor{#1}}
\newcommand{\AF}[1]{\AgdaFunction{#1}}
\newcommand{\AFi}[1]{\AgdaField{#1}}
\newcommand{\AB}[1]{\AgdaBound{#1}}
%\newcommand{\full}[1]{} %for the conference version
\newcommand{\full}[1]{#1} % for the full version

% New commands for Agda inline code
\newcommand{\D}{\AgdaDatatype}
\newcommand{\IC}{\AgdaInductiveConstructor}
\newcommand{\F}{\AgdaFunction}
\newcommand{\Ar}{\AgdaArgument}
\newcommand{\Fi}{\AgdaField}

\newcommand{\remove}[1]{}

\newcommand{\ie}{\emph{i.e.,} }
\newcommand{\cf}{\emph{c.f.} }
\newcommand{\eg}{\emph{e.g.,} }
\newcommand{\etal}{\emph{et~al.}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\wrt}{w.r.t.~}
\newcommand{\resp}{resp.~}
\newcommand{\loccit}{\emph{loc. cit.} }

\newcommand{\GTT}{\textbf{GTT}}
\newcommand{\AM}{\textbf{AM}}
\newcommand{\Set}{\mathbf{Set}}

%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
% for section-numbered lemmas etc., use "numberwithinsect"

\usepackage{microtype}%if unwanted, comment out or use option "draft"

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\usepackage{cleveref}

\bibliographystyle{plainurl}% the mandatory bibstyle

\newcommand{\NV}[1]{\textcolor{green}{NV: #1}}
\newcommand{\NW}[1]{\textcolor{orange}{NW: #1}}

\begin{document}

\maketitle

\begin{abstract}
  In type theory, programming and reasoning with possibly
  non-terminating programs and potentially infinite objects is
  achieved using coinductive types. Recursively defined programs of
  these types need to be productive to guarantee the
  consistency of the type system. Proof assistants such as Agda and
  Coq traditionally employ strict syntactic productivity checks, which
  often make programming with coinductive types convoluted.  One
  way to overcome this issue is by encoding productivity at the level
  of types so that the type system forbids the implementation of
  non-productive corecursive programs.
%%, meaning
%%  that although their definitions can be unfolded \emph{ad infinitum},
%%  finite unfoldings are guaranteed to produce . 
  In this paper we compare two different approaches to type-based
  productivity: guarded recursion and sized types. More specifically,
  we show how to simulate guarded recursion in Agda using sized
  types. We formalize the syntax of a simple type theory for guarded
  recursion, which is a variant of Atkey and McBride's
  calculus for productive coprogramming. Then we give a denotational semantics
  using presheaves over the preorder of sizes. Sized types are
  fundamentally used to interpret the characteristic
  features of guarded recursion, notably the fixpoint combinator.
%Sized types are fundamentally employed in the interpretation of the characteristic features of guarded recursion, notably the fixpoint combinator.
%%extending type theory with
%%  coinduction and coinductive types.  More specifically, we show that
%%  guarded recursion and guarded recursive types can be simulated with
%%  sized types.  We do this by defining a model of guarded recursion in
%%  type theory extended with sized types.  The model is variation of
%%  the topos of trees, and it is formalized in Agda.
\end{abstract}

\section{Introduction}
\label{sec:intro}
\input{Introduction}
%\nocite{*}

\section{The Host Type Theory}
\label{sec:prelim}
\input{Prelude/Basics}
%% \input{Presheaves/Presheaves}
%% \input{Presheaves/Const}
%% \input{Presheaves/Terminal}
%% \input{Presheaves/Product}
%% \input{Presheaves/Sum}
%% \input{Presheaves/Exp}
%% \input{CloTT/Structure/Contexts}
%% \input{CloTT/Structure/Types}
%% \input{CloTT/Structure/Terms}
%% \input{CloTT/Structure/DefinitionalEquality}
%% \input{CloTT/Structure/ContextOperations}
%% \input{CloTT/Structure/Subst}
%% \input{CloTT/TypeFormers/UnitType}
%% \input{CloTT/TypeFormers/SumType}
%% \input{CloTT/TypeFormers/ProductType}
%% \input{CloTT/TypeFormers/FunctionType}
%% \input{CloTT/TypeFormers/ClockQuantification}
%% \input{CloTT/TypeFormers/Later}
%% \input{CloTT/TypeFormers/LaterType}
%% \input{CloTT/TypeFormers/WeakenClock}
%% \input{CloTT/TypeFormers/Fix}
%% \input{CloTT/TypeFormers/Force}

%% \input{allcode}
%\subsection{Sized Types}

\section{The Object Type Theory}
\label{sec:syntax}
\input{Prelude/Syntax}
\input{Streams}

\section{Categorical Semantics}
\label{sec:presheaf_sem}

Next we give a categorical semantics for the calculus introduced in
\Cref{sec:syntax}.  We take inspiration from M{\o}gelberg's model
\cite{Mogelberg14} in which a (simple) type in a clock context
containing $n$ clocks is interpreted
as a presheaf in $\Set^{(\omega^n)^{\text{op}}}$, where $\omega$ is
the preorder of ordered natural numbers. In our model, we replace the
category $\omega$ with the preorder of sizes. Moreover, we only
consider the cases where either the clock context is empty or it
contains exactly one clock. This means that a type in \GTT\ is either
interpreted as an element of \F{Set} or as a presheaf over \F{Size}. 
In this section, we show how to implement presheaves and how to model
the fragment of \GTT\ corresponding to simply typed lambda
calculus. The interpretation of the guarded recursive and coinductive
features of \GTT\ is given in Section \ref{sec:guarded}.
%% interpreted a $n = 0$ or $n = 1$. where Since in \GTT\
%% every 
%% consider clock contexts containing at most one clock.
%% The type
%% \AD{ClockCtx}  of syntactic clock contexts is also used for semantic clock contexts.
%% 
%% 
%% 
%% a variation of the topos of
%% trees, where as the indexing category we replace the category of
%% ordered natural numbers with the preorder of sizes.
%% 
%% he similar to To do so, we use a similar method as used by M{\o}gelberg \cite{Mogelberg14}.
%% The interpretation of types depend on the clock contexts.
%% If the clock context is empty, then types become sets.
%% If the clock context has one clock, then types are interpreted as objects in the topos of trees.

\subsection{Presheaves}
\label{sec:presheaves}
\input{Presheaves/Presheaves}

\input{Presheaves/NaturalTransformations}

\input{Presheaves/Product}
The weakening type former \IC{↑} of \GTT\ is modeled using the constant presheaf, which we denote by \AF{Const}.
\input{Presheaves/Exp}

\subsection{Modelling Simple Types}
\label{sec:kripke}
\input{CloTT/Structure/Contexts}
\input{CloTT/Structure/Types}
\input{CloTT/Structure/Terms}
\input{CloTT/Structure/Subst}
%\input{CloTT/Structure/DefinitionalEquality}

We also need to provide a semantic version of 
%To interpret simple type theory, we also need to give 
the context operations, the simple type formers and the operations on substitutions.
Since their definitions are standard, we do not discuss them in detail.
For each of them, we need to make a case distinction based on the clock context.
For example, the empty variable context \IC{•} is interpreted
as the unit type in the clock context \IC{∅}, and it is interpreted as 
the terminal presheaf in the clock context \IC{κ}.
%, depending on the state of the clock context.
We use the operations on presheaves defined in \Cref{sec:presheaves} to interpret simple type formers.
\input{CloTT/TypeFormers/FunctionType}

\section{Modelling Guarded Recursion}
\label{sec:guarded}
In this section, we add the required guarded
recursive and coinductive features to the denotational semantics.
%Now we have the simple type formers and their terms.
%The next step is to add the type formers required for guarded recursion.
We start by defining the semantic box modality together with its
introduction and elimination rule. Then we construct the semantic
later modality. We show how to define the fixpoint combinator and the
force operation using sized types. In the end, we discuss how to model
guarded recursive types.
%% introduce the
%% semantic 
%%  together with the
%% terms associated 
%% and how to construct terms of those types.
%% With this in place, we define the fixpoint combinator.
%% In the end, we give the type former for guarded recursive types.

\subsection{Context Weakening and the Box Modality}
\input{CloTT/TypeFormers/WeakenClock}
\input{CloTT/TypeFormers/ClockQuantification}

\subsection{The Later Modality}
\label{sec:later}
\input{CloTT/TypeFormers/Later}
\input{CloTT/TypeFormers/Fix}
\input{CloTT/TypeFormers/Force}

\subsection{Guarded Recursive Types}
\label{sec:grt}
\input{CloTT/TypeFormers/Mu}

\section{Soundness and Consistency}
\label{sec:consistent}
\input{Prelude/Interpretation}
\input{CloTT/InterpretSyntax}
\input{GuardedTT}

%\section{Interpretation and Consistency}
%\input{GuardedTTAlt}

\section{Conclusions and Future Work}
\label{sec:concl}
We presented a simple type theory for guarded recursion that we called \GTT. We
formalized its syntax and semantics in Agda. Sized types
were employed to interpret the characteristic
features of guarded recursion. From this, we conclude that guarded
recursion can be simulated using sized types. We greatly benefited
from the fact that sized types constitute a native feature of Agda, so
we were able to fully develop our theory inside a proof assistant. 

This work can be extended in several different directions.
Various type theories for guarded recursion in the literature include
dependent types. Currently, there exist two disciplines for mixing
dependent types with guarded recursion: delayed substitutions
\cite{bizjak2016guarded} and ticks \cite{BahrGM17}. It would be interesting to extend
the syntax and semantics of \GTT\ with dependent types following one
of these two methods.

Abel and Vezzosi \cite{AbelV14} formalized a simple type theory
extended with the later modality in Agda. They focus on operational
properties of the calculus and give a certified proof of strong
normalization. As future work, we plan to investigate the metatheory
and the dynamic behavior of \GTT\ by formalizing a  type
checker and a normalization algorithm. This would also set up the
basis for a usable implementation of \GTT.



\remove{
and weakening, and we constructed denotational semantics for it in type theory extended with sized types.
From this, we concluded that this theory is consistent and that guarded recursion can be defined via sized types.

There are several possible ways to improve on this result.
First of all, one could define a dependent version of this type theory extended with multiple clocks.
This can be done in two ways.
Either one extends the object theory with delayed substitutions \cite{bizjak2016guarded} or with ticks \cite{BahrGM17}.
Second of all, with the deep embedding in place, one can further study the meta-theory by defining a formalized type checking and a normalization algorithm.
}

\remove{
Finally, we want to understand whether it is possible to reduce sized
types to guarded recursion.  This does not seems directly doable,
since sizes are not uniquely employed with coinductive types as a tool
for easing the productivity check of corecursive definitions, but they
are also used as auxiliary arguments of inductive types for
facilitating the termination check of recursive definitions. It seems
that, in order to capture this dual aspect of sizes, one needs to
extend the existing type theories for guarded recursion with
additional features.
}

In \GTT\, we restrict the least fixpoint 
operator \IC{μ} to act exclusively on strictly positive functors. This restriction is already 
present in Atkey and McBride's calculus, which is aimed at 
encoding coinductive types using the later modality and it does 
not allow solving general guarded recursive domain equations.
% such as\Ar{D} = ▻ \Ar{D → D}.
\GTT\ is a variant of Atkey and McBride's type theory, therefore the
similar restriction to stricly positive functors.  From the
formalization perspective, this restriction allows us to use Agda's
inductive types to model the guarded recursive types of GTT, as shown
in Section \ref{sec:grt}. In future work, we plan to extend \GTT\ with
the possibility of solving general guarded recursive domain
equations. This could be done in two ways: extending the language with
a universe, which allows the encoding of guarded recursive types as
fixpoints in the universe as shown e.g. by M{\o}gelberg
\cite{Mogelberg14}; or considering a \IC{μ} type former which, besides
stricly positive functors, also operates on functors where all
variables are guarded by an occurrence of the later modality. In the
second option we have to consider strictly positive functors to encode
usual inductive types, such as the natural numbers.

Finally, we would like to understand whether sized types can be
simulated using guarded recursion.
This could either be done by modeling sized types in a topos of
trees-like category \cite{BMSS-synthetic,MannaaM18} or via the
encoding of a type theory with sized types into a dependent type
theory for guarded recursion such as Clocked Type Theory
\cite{BahrGM17}. 
%Sized types require dependent types and subtyping, which \GTT\ does not have.
%We conjecture that Clocked Type Theory \cite{BahrGM17} can model sized types with explicit subtyping.


%%
%% Bibliography
%%

%% Please use bibtex, 

\bibliography{literature}

\input{Appendix}

\end{document}

